---
title: "TS Linear Model"
author: "Armando Torres Bejar"
header-includes:
   - \usepackage[spanish]{babel}
   - \usepackage[utf8]{inputenc}
   - \decimalpoint
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=2)
```
```{r include=FALSE}
library("openxlsx")
library("readxl")
library("zoo")
library("tidyverse")
library("dplyr")
library("graphics")
library("ggplot2")
library("foreign")
library("moments")
library("haven")
library("lmtest")
library("dynlm")
library("tseries")
library("urca")
library("Hmisc")
library("vars")
library("carData")
library("car")

ACT_IND_MEX <- read_excel("~/DS/Base_Datos_Dipl_ModXII_2024_LS.xlsx", sheet = "act_ind_mex")
```

Fist of all, we need to transform all of our variables that we´re going to use into time series variables for accurtae outputs
```{r}
attach(ACT_IND_MEX)

Periodo <- ts(Periodo,frequency = 4, start = c(1993,1))
ai <- ts(ai,frequency = 4, start = c(1993,1))
y <- ts(y,frequency = 4, start = c(1993,1), end = c(2024,2))
inf <- ts(inf,frequency = 4, start = c(1993,1))
ied <- ts(ied,frequency = 4, start = c(1993,1))
r <- ts(r,frequency = 4, start = c(1993,1))
```

#### Descriptive Analysis

Variables:

- ai = Industrial Activity
- y = Gross Domestic Product
- inf = Inflation Rate
- ied = Foreign Direct Investment
- r = Interest Rate (CETES 91 days): Average Profit

```{r include=FALSE}
vars <- c("ai","y","inf","ied","r")

mean <-c(
mean(ai),
mean(y),
mean(inf),
mean(ied),
mean(r)
)

variation <- c(
round(var(ai),digits = 2),
round(var(y),digits = 2),
round(var(inf),digits = 2),
round(var(ied),digits = 2),
round(var(r),digits = 2)
)

sd <- c(
sd(ai),
sd(y),
sd(inf),
sd(ied),
sd(r)
)

skew <- c(
skewness(ai, na.rm = FALSE),
skewness(y, na.rm = FALSE),
skewness(inf, na.rm = FALSE),
skewness(ied, na.rm = FALSE),
skewness(r, na.rm = FALSE)
)

kurtos <- c(
kurtosis(ai, na.rm = FALSE),
kurtosis(y, na.rm = FALSE),
kurtosis(inf, na.rm = FALSE),
kurtosis(ied, na.rm = FALSE),
kurtosis(r, na.rm = FALSE)
)

```

```{r}
dat <- data.frame(vars, mean, variation, skew, kurtos)
dat[,2:5] <- round(dat[,2:5], digits = 2)
dat

cor(ACT_IND_MEX[,c("ai","inf", "r")])
```

As the kurtosis is closer to 0, less skewness the variable has, in other words, a krtosis equal to 0 inidicates a normal distribution

#### Including Plots

```{r pressure, echo=FALSE}
par(mfrow=c(1,3))

ts.plot(ai, main="Industrial Activity: 1993-2024")
#ts.plot(y, main="PIB en Mexico: 1993-2024")
ts.plot(inf, main="Inflation Rate: 1993-2024")
#ts.plot(ied, main="Inversión extranjera a Mexico: 1993-2024")
ts.plot(r, main="Interest Rate: 1993-2024")

hist(ai, main="Industrial Activity")
#hist(y, main="Histograma del PIB en México")
hist(inf, main="Inflation Rate")
#hist(ied, main="Histograma de la Inversión extranjera a Mexico")
hist(r, main="Interest Rate")
```

As a first step, we´d like to know if the series is stationary, which guarantees that the statistical properties in the time series don´t change over time.<br> For this, we´ll be evaluating each of the variables with the Augmented Dickey Fuller Test, this test evaluates if the time series has a unit root.<br> The presence of a unit root implies that the time series is non-stationary.<br>
$$H_0: Non-stationary$$
The results interpretation focus in the Test statistical value compared to the Critical values, if Test statistical value < Critical value, we can reject the null hypothesis.
<br>
<br>

#### Estimate ADF test on the variables:

We´ll be evalating three different types of ADF test on our variables:

- With Trend and Intercept:
  - Tests whether the series is stationary around a deterministic trend.
- With Intercept Only:
  - Tests whether the series is stationary around a constant mean.
- Without Trend and Intercept:
  - Tests whether the series is stationary around 0.

$$\text{ADF test on variable: } ai$$

Tests whether the series is stationary with trend and intercept:
```{r}
summary(ur.df(ai, type=c("trend"), lags = 8, selectlags=c("BIC")))
```
Non stationary due to the test result: -2.94 which isn't smaller than the critical values.

Tests whether the series is stationary with intercept only:
```{r}
summary(ur.df(ai, type=c("drift"), lags = 8, selectlags=c("BIC")))
```
Non stationary due to the test result: -2.11  which isn't smaller than the critical values.

Tests whether the series is stationary without intercept:
```{r}
summary(ur.df(ai, type=c("none"), lags = 8, selectlags=c("BIC")))
```
Non stationary due to the test result: 1.5002  which isn't smaller than the critical values.

At this point, due to the ADF test, we can not reject the null hypothesis of unit root in any test

$$\text{ADF test on variable: } y$$

Tests whether the series is stationary with trend and intercept:
```{r}
summary(ur.df(y, type=c("trend"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -3.45 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with trending on the model.

Tests whether the series is stationary with intercept only:
```{r}
summary(ur.df(y, type=c("drift"), lags = 8, selectlags=c("BIC")))
```
Non stationary due to the test result: -1.16  which isn't smaller than the critical values.

Tests whether the series is stationary without intercept:
```{r}
summary(ur.df(y, type=c("none"), lags = 8, selectlags=c("BIC")))
```
Non stationary due to the test result: 3.42  which isn't smaller than the critical values.

The series seems to be stationary when considering trend, <br>
However, when considering only a drift or no deterministic component (no constant or trend), the series seems to be non stationary.

$$\text{ADF test on variable: } inf$$
Tests whether the series is stationary with trend and intercept:
```{r}
summary(ur.df(inf, type=c("trend"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -3.91 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with trending on the model.

Tests whether the series is stationary with intercept only:
```{r}
summary(ur.df(inf, type=c("drift"), lags = 8, selectlags=c("BIC")))
```
Non stationary due to the test result: -2.42  which isn't smaller than the critical values.

Tests whether the series is stationary without intercept:
```{r}
summary(ur.df(inf, type=c("none"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -1.98 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable without intercept on the model.

The test suggests to use this variable with no transformation due to the results regarding to the ADF test, which demonstrates that the variable is stationary without intercept and with trend and intercept.

$$\text{ADF test on variable: } ied$$
Tests whether the series is stationary with trend and intercept:
```{r}
summary(ur.df(ied, type=c("trend"), lags = 8, selectlags=c("BIC")))
```
Non stationary due to the test result: -1.79  which isn't smaller than the critical values.

Tests whether the series is stationary with intercept only:
```{r}
summary(ur.df(ied, type=c("drift"), lags = 8, selectlags=c("BIC")))
```
Non stationary due to the test result: -2.46  which isn't smaller than the critical values.

Tests whether the series is stationary without intercept:
```{r}
summary(ur.df(ied, type=c("none"), lags = 8, selectlags=c("BIC")))
```
Non stationary due to the test result: 0.45  which isn't smaller than the critical values.

Due to the ADF test, we can not reject the null hypothesis of unit root in any test

$$\text{ADF test on variable: } r$$
Tests whether the series is stationary with trend and intercept:
```{r}
summary(ur.df(r, type=c("trend"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -3.91 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with trending on the model.

Tests whether the series is stationary with intercept only:
```{r}
summary(ur.df(r, type=c("drift"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -3.91 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with intercept only on the model.

Tests whether the series is stationary without intercept:
```{r}
summary(ur.df(r, type=c("none"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -1.98 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable without intercept on the model.

We can reject the null hypothesis of unit root.

The time series is non stationary on all variables, the tests suggest differencing some of the vaiables in the series, which means substracting the previous value from the current value.<br> <br>¿Why differencing the time series?<br>Usually differencing the variables in the series helps removing trends, stabilizing the mean and removing seasonality.<br> Differencing is one of the transformations we can apply to the series, other options include: Log Transformation, De-Trending and De-Seasonalizing.

Differentiate the suggested variables by the previous ADF tests:
```{r echo=TRUE}
dai <- diff(ai)
dy <- diff(y)
died <- diff(ied)
```
#### Estimate ADF test on the differentiated variables:

$$\text{ADF test on variable: } dai$$
Tests whether the series is stationary with trend and intercept:
```{r}
summary(ur.df(dai, type=c("trend"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -3.91 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with trending on the model.

Tests whether the series is stationary with intercept only:
```{r}
summary(ur.df(dai, type=c("drift"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -3.91 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with intercept only on the model.

Tests whether the series is stationary without intercept:
```{r}
summary(ur.df(dai, type=c("none"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -1.98 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable without intercept on the model.

We can reject the null hypothesis of unit root.

$$\text{ADF test on variable: } dy$$
Tests whether the series is stationary with trend and intercept:
```{r}
summary(ur.df(dy, type=c("trend"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -9.84 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with trending on the model.

Tests whether the series is stationary with intercept only:
```{r}
summary(ur.df(dy, type=c("drift"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -9.85 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with intercept only on the model.

Tests whether the series is stationary without intercept:
```{r}
summary(ur.df(dy, type=c("none"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -9.22 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable without intercept on the model.

We can reject the null hypothesis of unit root.

$$\text{ADF test on variable: } died$$

Tests whether the series is stationary with trend and intercept:
```{r}
summary(ur.df(dy, type=c("trend"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -8.18 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with trending on the model.

Tests whether the series is stationary with intercept only:
```{r}
summary(ur.df(dy, type=c("drift"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -8.2 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable with intercept only on the model.

Tests whether the series is stationary without intercept:
```{r}
summary(ur.df(dy, type=c("none"), lags = 8, selectlags=c("BIC")))
```
Stationary, we can reject the null hypothesis because the test result: -8.12 is smaller than the than the 5pct critical value. Therefore, this is a stationary variable without intercept on the model.

We can reject the null hypothesis of unit root.